---
title: "Assignment 3: Air Quality Risk Assessment"
author: "Grace Bryan & Elena Nirgiotis"
date: "May 13, 2026"
format: 
  html:
    theme: cosmo
    toc: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## **Load Packages**

```{r}
library(ggplot2)
```

## **Part 1: Build the Dataset**

```{r}
# create a sequence of PM2.5 values to sample from
possible_pm25 <- seq(from = 2, to = 50, by = 0.1)

# sample 365 days worth of readings 
pm25_readings <- sample(possible_pm25, size = 365, replace = TRUE)

# make it a data frame 
air_quality <- data.frame(day = 1:365,
                          pm25 = pm25_readings)

# take a look
head(air_quality)

summary(air_quality$pm25)

# assign season based on day number
# days 1-90 = winter, 91-180 = spring, 181-270 = summer, 271-365 = fall
air_quality$season <- ifelse(air_quality$day <= 90, "winter",
                      ifelse(air_quality$day <= 180, "spring",
                      ifelse(air_quality$day <= 270, "summer", "fall")))

# store it as a factor 
air_quality$season <- factor(air_quality$season, 
                             levels = c("winter", "spring", "summer", "fall"))
# check it looks right
head(air_quality)
tail(air_quality)

summary(air_quality$season)

# add risk column, initialized to NA for now
air_quality$risk <- NA

# check it
head(air_quality)

summary(air_quality)


```

### **Why We Chose This Distribution**

We used `sample()` to randomly pick 365 PM2.5 values from a range of 2 to 50 µg/m³. We chose this range because we looked up typical urban PM2.5 levels and according to the EPA, most days in a city fall somewhere between 5 and 25 µg/m³, but it can spike higher during things like wildfires or heavy traffic. We thought a range of 2 to 50 made sense because it covers normal days but also includes some realistically bad ones, so we'd see a mix of all three risk categories in the data.

## **Part 2: Classify Risk with a For Loop**

### **Risk Classification Rules**

-   **good:** PM2.5 below 12 µg/m³
-   **moderate:** PM2.5 between 12 and 35.4 µg/m³
-   **unhealthy:** PM2.5 above 35.4 µg/m³

### **Use a For Loop to Fill Risk Column**

```{r}

current_streak <- 0
max_unhealthy_streak <- 0

for (i in 1:nrow(air_quality)) {
  
  # add today's pm25
  today_pm25 <- air_quality$pm25[i]
  
  # classify risk
  if (today_pm25 < 12) {
    air_quality$risk[i] <- "good"
    
  } else if (today_pm25 <= 35.4) {
    air_quality$risk[i] <- "moderate"
    
  } else {
    air_quality$risk[i] <- "unhealthy"
  }
  
  # check the streak - was yesterday also unhealthy?
  if (air_quality$risk[i] == "unhealthy") {
    
    if (i > 1 && air_quality$risk[i-1] == "unhealthy") {
      current_streak <- current_streak + 1  # keep the streak going
    } else {
      current_streak <- 1  # start a fresh streak
    }
    
  } else {
    current_streak <- 0  # not unhealthy so reset
  }
  
  # update max if current streak is the longest so far
  if (current_streak > max_unhealthy_streak) {
    max_unhealthy_streak <- current_streak
  }
  
}

# check results
table(air_quality$risk)
cat("Longest unhealthy streak:", max_unhealthy_streak, "days")

```

## **Part 3: Seasonal Summary with a For Loop**

### **Count Unhealthy Days by Season**

```{r}

# create a vector to store our counts, one slot per season
unhealthy_counts = c(winter = 0, spring = 0, summer = 0, fall = 0)

# loop through every day
for (i in 1:nrow(air_quality)) {
  
  # use today's season and risk
  today_season = air_quality$season[i]
  today_risk = air_quality$risk[i]
  
  # if today is unhealthy, add 1 to that season's count
  if (today_risk == "unhealthy") {
    unhealthy_counts[today_season] = unhealthy_counts[today_season] + 1
  }
  
}

# check results
unhealthy_counts

```

### **Convert Seasonal Summary to Data Frame**

```{r}

# convert to data frame for plotting
unhealthy_summary = data.frame(
  season = names(unhealthy_counts),
  unhealthy_days = as.numeric(unhealthy_counts))

unhealthy_summary

```

## **Part 4: Visualize and Report**

### **Plot 1: PM2.5 Across the Year**

```{r}

# make the title first using sprintf, including our max streak
plot_title = sprintf("Daily PM2.5 Concentrations Across the Year - Longest Unhealthy Streak: %d days", max_unhealthy_streak)

# now plot
plot_title = sprintf("Daily PM2.5 Concentrations Across the Year - Longest Unhealthy Streak: %d days", max_unhealthy_streak)

ggplot(air_quality, aes(x = day, y = pm25, color = season)) +
  geom_line() +
  labs(title = plot_title,
       x = "Day of Year",
       y = "PM2.5 (ug/m3)") +
  theme_minimal()
```

### **Plot 2: Count of Risk Categories**

```{r}

# make the title using sprintf
bar_title = sprintf("Count of Days by Risk Category - Longest Unhealthy Streak: %d days", max_unhealthy_streak)

# now plot
ggplot(air_quality, aes(x = risk, fill = risk)) +
  geom_bar() +
  labs(title = bar_title,
       x = "Risk Category",
       y = "Number of Days") +
  theme_minimal()
```

## **Management Summary**

This data is a simulation of 365 days of PM2.5 air quality readings over the course of a year. The majority of days fell into the moderate risk category, with a smaller number of good days (around 75). There were around 110 unhealthy days, and the longest consecutive streak of unhealthy days was 5 days in a row. Out of all the seasons, fall had the most unhealthy days (39), which may be due to wildfires that occur that time of year. From a management perspective, launching a public alert system that automatically notifies residents when unhealthy conditions are forecast, particularly heading into fall could be useful in ensuring public health. Additionally, we recommend investigating the specific emission sources contributing to the high number of unhealthy days and considering whether stricter local regulations or emission controls are warranted. Lastly, continued monitoring is essential to track whether conditions improve over time.
