---
title: "Assignment 3: Air Quality Risk Assessment"
author: "Grace Bryan & Elena Nirgiotis"
date: "May 13, 2026"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## **Overview**

## **Load Packages**

```{r}






```

## **Part 1: Build the Dataset**

```{r}
# create a sequence of PM2.5 values to sample from
possible_pm25 <- seq(from = 2, to = 50, by = 0.1)

# sample 365 days worth of readings 
pm25_readings <- sample(possible_pm25, size = 365, replace = TRUE)

# make it a data frame 
air_quality <- data.frame(day = 1:365,
                          pm25 = pm25_readings)

# take a look
head(air_quality)

summary(air_quality$pm25)

# assign season based on day number
# days 1-90 = winter, 91-180 = spring, 181-270 = summer, 271-365 = fall
air_quality$season <- ifelse(air_quality$day <= 90, "winter",
                      ifelse(air_quality$day <= 180, "spring",
                      ifelse(air_quality$day <= 270, "summer", "fall")))

# store it as a factor 
air_quality$season <- factor(air_quality$season, 
                             levels = c("winter", "spring", "summer", "fall"))
# check it looks right
head(air_quality)
tail(air_quality)

summary(air_quality$season)

# add risk column, initialized to NA for now
air_quality$risk <- NA

# check it
head(air_quality)

```

### **Why We Chose This Distribution**

text

### **Create the Dataset**

```{r}

```

## **Part 2: Classify Risk with a For Loop**

### **Risk Classification Rules**

-   **good:** PM2.5 below 12 µg/m³
-   **moderate:** PM2.5 between 12 and 35.4 µg/m³
-   **unhealthy:** PM2.5 above 35.4 µg/m³

### **Use a For Loop to Fill Risk Column**

```{r}

```

## **Part 3: Seasonal Summary with a For Loop**

### **Count Unhealthy Days by Season**

```{r}

```

### **Convert Seasonal Summary to Data Frame**

```{r}

```

## **Part 4: Visualize and Report**

### **Plot 1: PM2.5 Across the Year**

```{r}

```

### **Plot 2: Count of Risk Categories**

```{r}

```

## **Management Summary**

text
