---
title: "Assignment 3: Air Quality Risk Assessment"
author: "Grace Bryan & Elena Nirgiotis"
date: "May 13, 2026"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## **Overview**

text

## **Load Packages**

```{r}
library(tidyverse)
```

## **Part 1: Build the Dataset**

### **Why We Chose This Distribution**

text

### **Create the Dataset**

```{r}
set.seed(262)

# create daily air quality dataset
air_quality <- data.frame(
  day = 1:365,
  pm2.5= rnorm(365, mean = 22, sd = 10)
)

# prevent negative PM2.5 values
air_quality$pm2.5 <- pmax(air_quality$pm2.5, 0)

# assign seasons based on day number
air_quality$season <- NA

air_quality$season[1:90] <- "winter"
air_quality$season[91:180] <- "spring"
air_quality$season[181:270] <- "summer"
air_quality$season[271:365] <- "fall"

# convert to factor
air_quality$season <- factor(
  air_quality$season,
  levels = c("winter", "spring", "summer", "fall")
)

# initialize empty risk column
air_quality$risk <- NA

```

## **Part 2: Classify Risk with a For Loop**

### **Risk Classification Rules**

- **good:** PM2.5 below 12 µg/m³
- **moderate:** PM2.5 between 12 and 35.4 µg/m³
- **unhealthy:** PM2.5 above 35.4 µg/m³

### **Use a For Loop to Fill Risk Column**

```{r}
# initialize streak trackers
current_streak <- 0
max_unhealthy_streak <- 0

# loop through each day
for (i in 1:nrow(air_quality)) {
  
  # assign risk category
  if (air_quality$pm2.5[i] < 12) {
    air_quality$risk[i] <- "good"
    current_streak <- 0
    
  } else if (air_quality$pm2.5[i] <= 35.4) {
    air_quality$risk[i] <- "moderate"
    current_streak <- 0
    
  } else {
    air_quality$risk[i] <- "unhealthy"
    
    # increase unhealthy streak
    current_streak <- current_streak + 1
    
    # update maximum streak if needed
    if(current_streak > max_unhealthy_streak) {
      max_unhealthy_streak <- current_streak
    }
  }
}
```

## **Part 3: Seasonal Summary with a For Loop**

### **Count Unhealthy Days by Season**

```{r}
# create named vector to store counts
unhealthy_counts <- c(
  winter = 0,
  spring = 0,
  summer = 0,
  fall = 0
)

# loop through seasons
for (season_name in levels(air_quality$season)) {
  
  # count unhealthy days for each season
  unhealthy_counts[season_name] <- sum(
    air_quality$season == season_name &
      air_quality$risk == "unhealthy"
  )
  
}

# results
unhealthy_counts
```

### **Convert Seasonal Summary to Data Frame**

```{r}
season_summary <- data.frame(
  season = names(unhealthy_counts),
  unhealthy_days = as.numeric(unhealthy_counts)
)

season_summary
```

## **Part 4: Visualize and Report**

### **Plot 1: PM2.5 Across the Year**

```{r}
plot_title_1 <- sprintf(
  "Daily PM2.5 Concentrations Across the Year (Max Unhealthy Streak = %s Days)",
  max_unhealthy_streak
)

# time series plot
ggplot(air_quality, aes(x = day, y = pm2.5, color = season)) +
  geom_line() +
  labs(
    title = plot_title_1,
    x = "Day of Year",
    y = expression("PM2.5 (" * mu * "g/m"^3 * ")")
  ) +
  theme_minimal()
```

### **Plot 2: Count of Risk Categories**

```{r}
plot_title_2 <- sprintf(
  "Air Quality Risk Category Counts (Max Unhealthy Streak = %s Days)",
  max_unhealthy_streak
)

# bar chart of risk categories

ggplot(air_quality, aes(x = risk, fill = risk)) +
  geom_bar() +
  labs(
    title = plot_title_2,
    x = "Risk Category",
    y = "Number of Days"
  ) +
  theme_minimal()
```

### **Management Summary**

text
